### 2012-10-01 doc/prelim/document/chapters/linear_problem/linear_problem.tex
spellcheck done. draft complete

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computatable
from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003}
as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

@@

From an iteration standpoint, as we choose $\boldsymbol{\delta}$ from
$\mathcal{K}$ and constrain it with $\mathcal{L}$, each iteration
performs a projection that systematically anihilates the components of
the residual that exists in $\mathcal{L}$. This then means that if our
convergence criteria for an iterative method is bound to the residual
of the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
projection step gaurantees us that the norm of the new residual will
never be worse than that of the previous step and will typically move
us towards convergence. Depending on the qualities of the system in
Eq~(\ref{eq:linear_problem}), the selection of the subspaces
$\mathcal{K}$ and $\mathcal{L}$ can serve to both gaurantee
convergence and optimize the rate at which the residual is decreased.

\subsection{Krylov Subspace Methods}
\label{subsec:krylov_methods}
Among the most common projection techniques used in practice are a
class of methods known as \textit{Krylov subspace methods}. Here, the
search subspace is defined as the \textit{Krylov subspace}:
\begin{equation}
  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}\:,
  \label{eq:krylov_subspace}
\end{equation}
where $m$ denotes the dimensionality of the subspace. In order to
accomodate a more general structure for the operator in
Eq~(\ref{eq:linear_problem}), we often choose an \textit{oblique}
projection method where $\mathcal{K} \neq \mathcal{L}$. If we choose
$\mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)$, then we are
ultimately solving the normal system $\ve{A}^T\ve{A}\ve{x} =
\ve{A}^T\ve{b}$ where $\ve{A}^T\ve{A}$ will be symmetric positive
definite if $\ve{A}$ is nonsingular, thereby expanding the range of
operators over which these methods are valid. This choice of
constraint subspace also then gives us the result via
Eq~(\ref{eq:linear_projection_constraint}) that the residual is
minimized for all $\boldsymbol{\delta} \in \mathcal{K}$, forming the
basis for the \textit{generalized minimum residual method} (GMRES)
\citep{saad_gmres:_1986}.

Choosing GMRES as our model Krylov method, we are first tasked with
finding a projector onto the subspace. We seek an orthonormal basis
for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
that is commonly based on, but not limited to, the \textit{Arnoldi}
recurrence relation. The Arnoldi procedure will generate an
orthonormal basis, $\ve{V}_m \in \mathbb{R}^{N \times m}$, via a
variant of the Gram-Schmidt procedure that re-applies the operator for
each consecutive vector, thus forming a basis that spans the subspace
in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
dimensionality, $m$, to that of the subspace, we will refer to such
recurrence relations as \textit{long recurrence relations}. Those
orthogonal projection procedures that have a dimensionality less than
$m$ will be refferred to as \textit{short recurrence relations}.  Once
$\ve{V}_m$ is found, per the constraint subspace definition it then
follows that its basis is defined as $\ve{W}_m = \ve{A}
\ve{V}_m$. Knowing the projections onto the search and constraint
subspaces, the GMRES iteration may be formulated as follows:
\begin{algorithm}[htpb!]
  \caption{GMRES Iteration}
  \label{alg:gmres}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
    \State $\beta := ||\ve{r}_0||_2$
    \State $\ve{v}_1 := \ve{r}_0 / \beta$
    \Comment{Create the orthonormal basis for the Krylov subspace}
    \For{$j = 1, 2, \cdots, m$}
    \State $h_{ij} \leftarrow \langle w_j,v_j \rangle$
    \State $w_j \leftarrow w_j - h_{ij}v_i$
    \EndFor
    \State $h_{j+1,j} \leftarrow ||w_j||_2$
    \State $v_{j+1} \leftarrow w_j / h_{j+1,j}$
    \Comment{Apply the orthogonality constraints}
    \State $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
    \ve{y}||_2 $
    \State $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
  \end{algorithmic}
\end{algorithm}

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_denovo:_2010})
that normally may not be amenable to its full construction. In
addition, the minimization problem is a straight-forward least-squares
problem where $\ve{H}$ is an upper-Hessenberg matrix.

\section{Parallel Projection Methods}
\label{sec:parallel_krylov_methods}
Modern parallel implementations of projection methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. For methods like GMRES, this arises from
the fact that Krylov methods require only a handful of operation types
in their implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_iterative_2003} and as
noted in Algorithm~\ref{alg:gmres}, these operations are
preconditioning, matrix-vector multiplications, vector updates, and
inner products. For the last three items, linear algebra libraries
such as PETSc \citep{gropp_scalable_1993} and Trilinos
\citep{heroux_overview_2005} provide efficient parallel
implementations for these operations. Depending on the type of
preconditioning used, efficient parallel implementations may also be
available for those operations. Due to their prevalence in modern
numerical methods, parallel formulations these operations have
warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
of scatter/gather operations are required such that global
communication operations must occur. Although the relative performance
of such operations is bound to the implementation, asymptotically
perfomance should be the same across all implementations.

@@

\subsection{Parallel Vector Update}
\label{subsec:parallel_vec_update}
Parallel vector update operations arise from the construction of the
orthonormal basis and the application of the correction generated by
the constraints to the solution vector. Vector update operations are
embarassingly parallel in that they require no communication
operations to be successfully completed; all data operated on is
local. These operations are globally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
  \:,
  \label{eq:global_vector_update}
\end{equation}
and locally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
  \:,
  \label{eq:local_vector_update}
\end{equation}
where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
size $N_l$, and $a \in \mathbb{R}^N$. In order to avoid communcation,
the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
decomposition where each parallel process owns the same pieces of each
vector.

@@

\ve{y}_1$. Here, some of the data is intrinisically local, and some

@@

\subsection{Parallel Perfomance Implications for Krylov Methods}
\label{subsec:projection_method_performance}
Knowing the parallel characteristics of the key operations we must
perform in order to implement Krylov methods, we can make a few
statements about parallel perfomance and implications for operation on
machines of increasing size.  For very large distributed machines,
global reduction operations required at several levels of Krylov
algorithms stand to reduce scalablitity and perfomance. Furthermore,
communication between adjacent domains in matrix-vector multiply
operations may also cause a bottleneck as the number of domains used
in simulation grows. The end result is that global data must be
collected and communicated. For scaling improvement, we seek a
reduction in these types of operations. In addition, these issues
become more prominent as the Krylov iterations progress, causing the
Krylov subspace to grow and the total number of operations needed to
orthogonalize that subspace to increase.

>>>>

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computable
from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003}
as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

@@

From an iteration standpoint, as we choose $\boldsymbol{\delta}$ from
$\mathcal{K}$ and constrain it with $\mathcal{L}$, each iteration
performs a projection that systematically annihilates the components of
the residual that exists in $\mathcal{L}$. This then means that if our
convergence criteria for an iterative method is bound to the residual
of the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
projection step guarantees us that the norm of the new residual will
never be worse than that of the previous step and will typically move
us towards convergence. Depending on the qualities of the system in
Eq~(\ref{eq:linear_problem}), the selection of the subspaces
$\mathcal{K}$ and $\mathcal{L}$ can serve to both guarantee
convergence and optimize the rate at which the residual is decreased.

\subsection{Krylov Subspace Methods}
\label{subsec:krylov_methods}
Among the most common projection techniques used in practice are a
class of methods known as \textit{Krylov subspace methods}. Here, the
search subspace is defined as the \textit{Krylov subspace}:
\begin{equation}
  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}\:,
  \label{eq:krylov_subspace}
\end{equation}
where $m$ denotes the dimensionality of the subspace. In order to
accommodate a more general structure for the operator in
Eq~(\ref{eq:linear_problem}), we often choose an \textit{oblique}
projection method where $\mathcal{K} \neq \mathcal{L}$. If we choose
$\mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)$, then we are
ultimately solving the normal system $\ve{A}^T\ve{A}\ve{x} =
\ve{A}^T\ve{b}$ where $\ve{A}^T\ve{A}$ will be symmetric positive
definite if $\ve{A}$ is nonsingular, thereby expanding the range of
operators over which these methods are valid. This choice of
constraint subspace also then gives us the result via
Eq~(\ref{eq:linear_projection_constraint}) that the residual is
minimized for all $\boldsymbol{\delta} \in \mathcal{K}$, forming the
basis for the \textit{generalized minimum residual method} (GMRES)
\citep{saad_gmres:_1986}.

Choosing GMRES as our model Krylov method, we are first tasked with
finding a projector onto the subspace. We seek an orthonormal basis
for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
that is commonly based on, but not limited to, the \textit{Arnoldi}
recurrence relation. The Arnoldi procedure will generate an
orthonormal basis, $\ve{V}_m \in \mathbb{R}^{N \times m}$, via a
variant of the Gram-Schmidt procedure that re-applies the operator for
each consecutive vector, thus forming a basis that spans the subspace
in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
dimensionality, $m$, to that of the subspace, we will refer to such
recurrence relations as \textit{long recurrence relations}. Those
orthogonal projection procedures that have a dimensionality less than
$m$ will be referred to as \textit{short recurrence relations}.  Once
$\ve{V}_m$ is found, per the constraint subspace definition it then
follows that its basis is defined as $\ve{W}_m = \ve{A}
\ve{V}_m$. Knowing the projections onto the search and constraint
subspaces, the GMRES iteration may be formulated as follows:
\begin{algorithm}[htpb!]
  \caption{GMRES Iteration}
  \label{alg:gmres}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
    \State $\beta := ||\ve{r}_0||_2$
    \State $\ve{v}_1 := \ve{r}_0 / \beta$
    \Comment{Create the orthonormal basis for the Krylov subspace}
    \For{$j = 1, 2, \cdots, m$}
    \State $h_{ij} \leftarrow \langle w_j,v_j \rangle$
    \State $w_j \leftarrow w_j - h_{ij}v_i$
    \EndFor
    \State $h_{j+1,j} \leftarrow ||w_j||_2$
    \State $v_{j+1} \leftarrow w_j / h_{j+1,j}$
    \Comment{Apply the orthogonality constraints}
    \State $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
    \ve{y}||_2 $
    \State $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
  \end{algorithmic}
\end{algorithm}

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potential to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_denovo:_2010})
that normally may not be amenable to its full construction. In
addition, the minimization problem is a straight-forward least-squares
problem where $\ve{H}$ is an upper-Hessenberg matrix.

\section{Parallel Projection Methods}
\label{sec:parallel_krylov_methods}
Modern parallel implementations of projection methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. For methods like GMRES, this arises from
the fact that Krylov methods require only a handful of operation types
in their implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_iterative_2003} and as
noted in Algorithm~\ref{alg:gmres}, these operations are
preconditioning, matrix-vector multiplications, vector updates, and
inner products. For the last three items, linear algebra libraries
such as PETSc \citep{gropp_scalable_1993} and Trilinos
\citep{heroux_overview_2005} provide efficient parallel
implementations for these operations. Depending on the type of
preconditioning used, efficient parallel implementations may also be
available for those operations. Due to their prevalence in modern
numerical methods, parallel formulations these operations have
warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
of scatter/gather operations are required such that global
communication operations must occur. Although the relative performance
of such operations is bound to the implementation, asymptotically
performance should be the same across all implementations.

@@

\subsection{Parallel Vector Update}
\label{subsec:parallel_vec_update}
Parallel vector update operations arise from the construction of the
orthonormal basis and the application of the correction generated by
the constraints to the solution vector. Vector update operations are
embarrassingly parallel in that they require no communication
operations to be successfully completed; all data operated on is
local. These operations are globally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
  \:,
  \label{eq:global_vector_update}
\end{equation}
and locally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
  \:,
  \label{eq:local_vector_update}
\end{equation}
where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
size $N_l$, and $a \in \mathbb{R}^N$. In order to avoid communication,
the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
decomposition where each parallel process owns the same pieces of each
vector.

@@

\ve{y}_1$. Here, some of the data is intrinsically local, and some

@@

\subsection{Parallel Performance Implications for Krylov Methods}
\label{subsec:projection_method_performance}
Knowing the parallel characteristics of the key operations we must
perform in order to implement Krylov methods, we can make a few
statements about parallel performance and implications for operation on
machines of increasing size.  For very large distributed machines,
global reduction operations required at several levels of Krylov
algorithms stand to reduce scalability and performance. Furthermore,
communication between adjacent domains in matrix-vector multiply
operations may also cause a bottleneck as the number of domains used
in simulation grows. The end result is that global data must be
collected and communicated. For scaling improvement, we seek a
reduction in these types of operations. In addition, these issues
become more prominent as the Krylov iterations progress, causing the
Krylov subspace to grow and the total number of operations needed to
orthogonalize that subspace to increase.
